RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-08-29 manuscripts/manuscript.tex
try simplified Dockerfile

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \item
    NeXML is a new phylogenetic informatics data and metadata exchange
    format that offers substantial improvements in reliability,
    extensibility, and richness over earlier standards such as NEXUS. In
    this paper we present the RNeXML package which provides a bridge
    between this rich and reliable data format and the extensive library
    of phylogenetic tools already available in R, with a particular
    emphasis on comparative methods.
  \item
    RNeXML is much richer than a basic parser: here we illustrate how the
    package facilitates adding and manipulating semantic annotations, that
    is annotations whose meaning can be defined externally by way of
    semantic web technologies.
  \item
    An essential feature of NeXML is to provide both the flexibility to be
    continually extended to meet the emerging needs of the phylogenetics
    community while remaining backwards-compatible with software developed
    for earlier versions. Here we illustrate how RNeXML makes it easy to
    extend NeXML in this manner.
  \item
    RNeXML is well suited for users to preserve and maintain their
    personal archives of phylogenetic and comparative trait data in
    consistent, query-able, and cross-platform format, to leverage the
    rapidly growing public community archives of rich NeXML files, and to
    interact with web services that deliver their results (with their
    provenance) as NeXML.
  \end{enumerate}
  \end{abstract}

@@

The R language (R Core Team 2014) has become an important environment
for phylogenetics research, with over 30 packages available including
several with over 1000 citations (`CRAN task view' 2014), though
stand-alone programs remain an important part of this ecosystem for
tasks such as generating phylogenetic trees from sequence data. Within
this ecosystem a variety of different data formats for phylogenetic data
have been developed and modified. Most but not all are based on the
simple and popular \texttt{ape::phylo} format provided by the ape
package (Paradis \emph{et al.} 2004).

@@

\texttt{RNeXML} opens up a rich world of both currently available and
rapidly growing data sets in the NeXML format to the R user and the
large ecosystem of R packages currently available for handling
phylogenetic data\\R packages and R users frequently add rich
information to the phylogenies they work with, but need a platform
independent data exchange format to share and archive this. We
illustrate how NeXML can be used to distribute both phylogenies and
character trait data used in many popular phylogenetic comparative
methods tools in R.\\\texttt{RNeXML} also provides convenient tools for
adding metadata to NeXML files, including the automated generation of
metadata and examples of how to extend the NeXML standard to represent
new data features without breaking validation or backwards compatibility
of the file type.

@@

NeXML can not only contain multiple phylogenetic trees in a single
document, but can also group sets of trees into separate blocks (e.g.~as
resulting from different workflows). The \texttt{get\_trees} function
attempts to return the most sensible R object for NeXML: if only a
single tree is available, it will return an \texttt{ape::phylo} object;
if multiple trees are available in a single \texttt{trees} block, it
will return an \texttt{ape::multiPhylo} object (a list of
\texttt{ape::phylo} objects). If multiple blocks are available, it will
return a list of \texttt{multiPhylo} blocks. While most users will find
this behavior convenient, there are times (such as looping over many
NeXML files) when having a variable return type is undesirable. To
always get back a list of \texttt{multiPhylo} objects, even when only
one tree is available, use \texttt{get\_trees\_list()}.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user. For instance,
a user can access all the phylogenies available in TreeBASE through the
\texttt{nexml} format. \texttt{RNeXML} can read directly from a URL:

@@

Previously, this was possible using the \texttt{treebase} package
{[}Boettiger \& {Temple Lang} (2012)``{]} for R, which relied on the
NEXUS parser provided in the \texttt{ape} package. Due to
inconsistencies present in the NEXUS standard (Vos \emph{et al.} 2012),
a significant number of TreeBASE NEXUS files could not be read in this
way. Using the more precisely defined NeXML standard, every valid NeXML
document on TreeBASE can now successfully be parsed to R objects.

@@

Like NEXUS files, NeXML documents can also contain character data. While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological)
data used to infer the phylogenetic tree, we find this ideal for
character data used in comparative methods popular in R. Methods to
extract character data work much like those for the phylogenetic data.
Here we read in an example NeXML file that contains both a phylogeny and
character data. \texttt{get\_characters} returns a \texttt{data.frame}
with columns as characters and rows as taxa:

@@

The first argument to \texttt{nexml\_write} can be a \texttt{ape::phylo}
phylogeny, \texttt{multiPhylo} list of phylogenies, or a \texttt{nexml}
object -- anything that can be coerced into the \texttt{nexml} class. As
we have seen, NeXML files can contain multiple phylogenetic trees, as
well as character matrices. We can add both trees and character data
using \texttt{nexml\_write},

@@

NeXML files can also be built up bit by bit using the
\texttt{add\_characters}, \texttt{add\_trees}, and \texttt{add\_meta}.
Each of these functions can add data to an existing \texttt{nexml}
object or create a new nexml object if none is provided. Here we first
create a \texttt{nexml} object containing the phylogeny data, and then
add the character data to it:

@@

To adjust these and other common metadata fields that should be part of
any data deposition, \texttt{RNeXML} provides the
\texttt{add\_basic\_meta()} function. This function does not assume any
familiarity with the concepts of namespaces or other more advanced
metadata tools discussed later. The \texttt{add\_basic\_meta} takes an
existing \texttt{nexml} object (or will create a new one if none is
provided, like the other \texttt{add\_} functions), and any of the
following fields: \texttt{title}, \texttt{description},
\texttt{creator}, \texttt{pubdate}, \texttt{rights}, \texttt{publisher},
\texttt{citation}. Additional fields may be added to subsequent
versions, see \texttt{?add\_basic\_metadata} for the most up-to-date
information. Here we create a new \texttt{nexml} object with the basic
metadata for the \texttt{bird.orders} data provided in the \texttt{ape}
package:

@@

The \texttt{taxize\_nexml} function uses the R package \texttt{taxize}
(Chamberlain \& Sz{ö}cs 2013) to check each taxon label against the NCBI
database. If a unique match is found, a metadata annotation is added to
the taxon providing the NCBI identification number to the taxonomic
unit. If no match is found, the user is warned to check for possible
typographic errors in the taxonomic labels provided.

@@

@@

From the list of namespaces in \texttt{prefixes}, we can follow links a
handful of established vocabularies and find terms that are not included
in \texttt{add\_basic\_meta}. For instance, by resolving the location
identified by the \texttt{prism} prefix,

@@

We can add this annotation to our existing \texttt{birds} NeXML file
using the \texttt{add\_meta} function. Because we do not specify a
level, it is added to the root node, referring to the NeXML file as a
whole.

@@

The \texttt{get\_trees} function can be used to return an
\texttt{ape::phylo} tree as usual. \texttt{RNeXML} automatically detects
the \texttt{simmap} reconstruction data and returns includes this in a
\texttt{maps} element of the \texttt{ape::phylo} object, for use with
other \texttt{phytools} functions.

@@

The \texttt{RNeXML} package is available through the CRAN repository
under a CC0 license. At the time of writing, the current version is
1.1.3. The current development branch can be found on Github at
{[}\url{https://github.com/ropensci/RNeXML}{]}. See the NEWS file for
changes in recent versions. Users are encouraged to submit bug reports
or feature requests in the issues log. Contact
\texttt{ropensci-discuss@googlegroups.com} or the package maintainer for
help, or email the phylogenetics R users group list at
\texttt{r-sig-phylo@r-project.org}.

There are many promising directions for further development of the
\texttt{RNeXML} package. The project's
\href{https://github.com/ropensci/RNeXML/issues/milestones}{Milestones
page} page provides a list of steps currently planned or under
consideration. In particular, these milestones include increased support
for interactive mode of entering data (wizards), conversions to and from
other existing R phylogenetics formats such those found in the
\texttt{phylobase} package (al. 2014), and support for integration of
semantics defined through the Web Ontology language (OWL), as discussed
in the Challenges section.

@@

al., R.H. et. (2014). \emph{phylobase: Base package for phylogenetic
structures and comparative data}. Retrieved from
\url{http://CRAN.R-project.org/package=phylobase}

@@

CRAN task view: Phylogenetics, especially comparative methods. (2014).
Retrieved from
\url{http://cran.r-project.org/web/views/Phylogenetics.html}

@@

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \item
    NeXML is a new phylogenetic informatics data and metadata exchange
    format that offers substantial improvements in reliability,
    extensibility, and richness over earlier formats such as NEXUS. In
    this paper we present the RNeXML package which provides a bridge
    between this rich and reliable data format and the extensive library
    of phylogenetic tools already available in R, with a particular
    emphasis on comparative methods.
  \item
    RNeXML is much richer than a basic parser: here we illustrate how the
    package facilitates adding and manipulating semantic annotations, that
    is annotations whose meaning can be defined externally by way of
    semantic web technologies.
  \item
    An essential feature of NeXML is to provide both the flexibility to be
    continually extended to meet the emerging needs of the phylogenetics
    community while remaining backwards-compatible with software developed
    for earlier versions. Here we illustrate how RNeXML makes it easy to
    extend NeXML in this manner.
  \item
    RNeXML is well suited for users to preserve and maintain their
    personal archives of phylogenetic and comparative trait data in
    consistent, query-able, and cross-platform format, to leverage the
    rapidly growing public community archives of rich NeXML files, and to
    interact with web services that deliver their results (with their
    provenance) as NeXML.
  \end{enumerate}
  \end{abstract}

@@

The R language (R Core Team 2014) has become an important environment
for phylogenetics research, with over 30 packages available including
several with over 1000 citations (O'Meara 2014), though stand-alone
programs remain an important part of this ecosystem for tasks such as
generating phylogenetic trees from sequence data. Within this ecosystem
a variety of different data formats for phylogenetic data have been
developed and modified. Most but not all are based on the simple and
popular \texttt{ape::phylo} format provided by the ape package (Paradis
\emph{et al.} 2004).

@@

\texttt{RNeXML} opens up a rich world of both currently available and
rapidly growing data sets in the NeXML format to the R user and the
large ecosystem of R packages currently available for handling
phylogenetic data.\\ R packages and R users frequently add rich
information to the phylogenies they work with, but need a platform
independent data exchange format to share and archive this. We
illustrate how NeXML can be used to distribute both phylogenies and
character trait data used in many popular phylogenetic comparative
methods tools in R.\\\texttt{RNeXML} also provides convenient tools for
adding metadata to NeXML files, including the automated generation of
metadata and examples of how to extend the NeXML standard to represent
new data features without breaking validation or backwards compatibility
of the file type.

@@

NeXML can not only contain multiple phylogenetic trees in a single
document, but can also group sets of trees into separate blocks (e.g.~as
resulting from different workflows). The \texttt{get\_trees()} function
attempts to return the most sensible R object for NeXML: if only a
single tree is available, it will return an \texttt{ape::phylo} object;
if multiple trees are available in a single \texttt{trees} block, it
will return an \texttt{ape::multiPhylo} object (a list of
\texttt{ape::phylo} objects). If multiple blocks are available, it will
return a list of \texttt{multiPhylo} blocks. While most users will find
this behavior convenient, there are times (such as looping over many
NeXML files) when having a variable return type is undesirable. To
always get back a list of \texttt{multiPhylo} objects, even when only
one tree is available, use \texttt{get\_trees\_list()}.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user. For instance,
a user can access all the phylogenies available in
\href{http://treebase.org/treebase-web/home.html}{TreeBASE} through the
\texttt{nexml} format. \texttt{RNeXML} can read directly from a URL:

@@

Previously, this was possible using the \texttt{treebase} package
(Boettiger \& {Temple Lang} 2012) for R, which relied on the NEXUS
parser provided in the \texttt{ape} package. Due to inconsistencies
present in the NEXUS standard (Vos \emph{et al.} 2012), a significant
number of TreeBASE NEXUS files could not be read in this way. Using the
more precisely defined NeXML standard, every valid NeXML document on
TreeBASE can now successfully be parsed to R objects.

@@

Like NEXUS files, NeXML documents can also contain character data. While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological)
data used to infer the phylogenetic tree, we find this ideal for
character data used in comparative methods popular in R. Methods to
extract character data work much like those for the phylogenetic data.
Here we read in an example NeXML file that contains both a phylogeny and
character data. \texttt{get\_characters()} returns a \texttt{data.frame}
with columns as characters and rows as taxa:

@@

The first argument to \texttt{nexml\_write()} can be a
\texttt{ape::phylo} phylogeny, \texttt{multiPhylo} list of phylogenies,
or a \texttt{nexml} object -- anything that can be coerced into the
\texttt{nexml} class. As we have seen, NeXML files can contain multiple
phylogenetic trees, as well as character matrices. We can add both trees
and character data using \texttt{nexml\_write()},

@@

NeXML files can also be built up bit by bit using
\texttt{add\_characters()}, \texttt{add\_trees()}, and
\texttt{add\_meta()}. Each of these functions can add data to an
existing \texttt{nexml} object or create a new nexml object if none is
provided. Here we first create a \texttt{nexml} object containing the
phylogeny data, and then add the character data to it:

@@

To adjust these and other common metadata fields that should be part of
any data deposition, \texttt{RNeXML} provides the
\texttt{add\_basic\_meta()} function. This function does not assume any
familiarity with the concepts of namespaces or other more advanced
metadata tools discussed later. The \texttt{add\_basic\_meta()} function
takes an existing \texttt{nexml} object (or will create a new one if
none is provided, like the other \texttt{add\_} functions), and any of
the following fields: \texttt{title}, \texttt{description},
\texttt{creator}, \texttt{pubdate}, \texttt{rights}, \texttt{publisher},
\texttt{citation}. Additional fields may be added to subsequent
versions, see \texttt{?add\_basic\_metadata} for the most up-to-date
information. Here we create a new \texttt{nexml} object with the basic
metadata for the \texttt{bird.orders} data provided in the \texttt{ape}
package:

@@

The \texttt{taxize\_nexml()} function uses the R package \texttt{taxize}
(Chamberlain \& Sz{ö}cs 2013) to check each taxon label against the NCBI
database. If a unique match is found, a metadata annotation is added to
the taxon providing the NCBI identification number to the taxonomic
unit.

@@

If no match is found, the user is warned to check for possible
typographic errors in the taxonomic labels provided. If multiple matches
are found, the user will be prompted to choose between them.

@@

From the list of namespaces in \texttt{prefixes}, we can follow links to
a handful of established vocabularies and find terms that are not
included in \texttt{add\_basic\_meta()}. For instance, by resolving the
location identified by the \texttt{prism} prefix,

@@

We can add this annotation to our existing \texttt{birds} NeXML file
using the \texttt{add\_meta()} function. Because we do not specify a
level, it is added to the root node, referring to the NeXML file as a
whole.

@@

The \texttt{get\_trees()} function can be used to return an
\texttt{ape::phylo} tree as usual. \texttt{RNeXML} automatically detects
the \texttt{simmap} reconstruction data and returns includes this in a
\texttt{maps} element of the \texttt{ape::phylo} object, for use with
other \texttt{phytools} functions.

@@

The \texttt{RNeXML} package is available through the CRAN repository
under a CC0 license. At the time of writing, the current version is
1.1.3. The current development branch can be found on Github at
\href{}{\url{https://github.com/ropensci/RNeXML}}. See the NEWS file for
changes in recent versions. Users are encouraged to submit bug reports
or feature requests in the issues log. Contact
\texttt{ropensci-discuss@googlegroups.com} or the package maintainer for
help, or email the phylogenetics R users group list at
\texttt{r-sig-phylo@r-project.org}.

There are many promising directions for further development of the
\texttt{RNeXML} package. The project's
\href{https://github.com/ropensci/RNeXML/issues/milestones}{Milestones
page} page provides a list of steps currently planned or under
consideration. In particular, these milestones include increased support
for interactive mode of entering data (wizards), conversions to and from
other existing R phylogenetics formats such those found in the
\texttt{phylobase} package (NESCENT R Hackathon Team 2014), and support
for integration of semantics defined through the Web Ontology language
(OWL), as discussed in the Challenges section.

@@

@@

@@

NESCENT R Hackathon Team. (2014). \emph{phylobase: Base package for
phylogenetic structures and comparative data}. Retrieved from
\url{http://CRAN.R-project.org/package=phylobase}

O'Meara, B. (2014). CRAN task view: Phylogenetics, especially
comparative methods. Retrieved from
\url{http://cran.r-project.org/web/views/Phylogenetics.html}
